Revealing transient strain in geodetic data with Gaussian process regression

### 2017-11-09 RevisedSubmission/hines_hetland_2017_gji.tex
did more

Transient strain rates derived from GNSS data can be used to detect
and understand geophysical phenomena such as slow slip events and
postseismic deformation. Here we propose using Gaussian process
regression (GPR) as a tool for estimating transient strain rates from
GNSS data. GPR is a non-parametric, Bayesian method for interpolating
scattered data. Transient strain rates estimated with GPR have
meaningful uncertainties, allowing geophysical signal to be easily
discerned from noise. In our approach, we assume a stochastic prior
model for transient displacements. The prior describes how much one
expects transient displacements to covary spatially and temporally. A
posterior estimate of transient strain rates is obtained by
differentiating the posterior displacements. One limitation with GPR
is that it is not robust against outliers, so we introduce a
pre-processing method for detecting and removing outliers in GNSS
data. As a demonstration, we use GPR to detect transient strain
resulting from slow slip events in the Pacific Northwest. Maximum
likelihood methods are used to constrain a prior model for transient
displacements in this region. The temporal covariance of our prior
model is described by a compact Wendland covariance function, which
significantly reduces the computational burden that can be associated
with GPR. Our results reveal the spatial and temporal evolution of
strain from slow slip events. We verify that the transient strain
estimated with GPR is in fact geophysical signal by comparing it to
the seismic tremor that is associated with Pacific Northwest slow slip
events.

@@

In addition to detecting geophysical processes, the GNSS derived
transient strain rates can be used to better understand the data from
borehole strain meters (BSMs). The Plate Boundary Observatory
maintains about forty BSMs in the Pacific Northwest, and it has been
demonstrated that BSMs are able to record transient geophysical events
such as SSEs \citep[e.g.,][]{Dragert2011}. However, there are
complications that prevent BSM data from being used quantitatively in
geophysical studies. One difficulty is that BSM data should be
calibrated with a well known strain source, such as diurnal and
semidiurnal tides \citep{Hart1996,Roeloffs2010,Hodgkinson2013}.
Unfortunately, the tidal forces at BSMs which record SSEs are strongly
influenced by local bodies of water such as the Straight of Juan de
Fuca, making it difficult to form a theoretical prediction of tidal
strains \citep{Roeloffs2010}. Another complication is that noise in
BSM data is not well understood. The noise consists, in part, of a
long-term decay resulting from the instrument equilibrating with the
surrounding rock \citep{Gladwin1987}. Typically, this noise is dealt
with in an ad-hoc manner by fitting and removing exponentials and
low-order polynomials. We envision that the GNSS derived strain rates
from this paper can be used as a reference strain for calibrating BSM
data and quantify its noise.

%% CONCLUSION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}\label{sec:Conclusion}
In this paper we propose using Gaussian process regression (GPR) to
estimate transient strain rates from GNSS data. Most other methods for
estimating strain rates assume a parametric representation of
deformation, which can bias the results if the parameterization is not
chosen carefully. Here we assume a stochastic, rather than parametric,
prior model for displacements. Our prior model describes how much we
expect transient displacements to covary spatially and temporally. If
we know nothing about the underlying signal that we are trying to
recover, then the prior model can be chosen objectively with maximum
likelihood methods. Because GPR is a Bayesian method, the
uncertainties on our estimated transient strain rates are well
quantified, allowing one to discern geophysical signal from noise. We
demonstrate that GPR is an effective tool for detecting geophysical
phenomena, such as slow slip events, in our application to GNSS data
from the Pacific Northwest. One limitation with GPR is that it is not
robust against outliers. To overcome this limitation, we have
introduced an effective pre-processing method for identifying and
removing outliers from GNSS datasets. Another complication with GPR is
that it usually involves inverting a dense matrix where the number of
rows and columns is equal to the number of observations. This is
prohibitive when using several years of daily GNSS observations from a
network of several hundred stations. We significantly reduce the
computational burden of GPR by using compact Wendland covariance
function to describe our prior model. While this paper just focuses on
estimating transient strain rates, we believe that GPR is a powerful
tool that can be applied to a wide range of geophysical problems.

>>>>

Transient strain rates derived from GNSS data can be used to detect
and understand geophysical processes such as slow slip events and
postseismic deformation. Here we propose using Gaussian process
regression (GPR) as a tool for estimating transient strain rates from
GNSS data. GPR is a non-parametric, Bayesian method for interpolating
scattered data. In our approach, we assume a stochastic prior model
for transient displacements. The prior describes how much we expect
transient displacements to covary spatially and temporally. A
posterior estimate of transient strain rates is obtained by
differentiating the posterior displacements, which are formed by
conditioning the prior with the GNSS data. As a demonstration, we use
GPR to detect transient strain resulting from slow slip events in the
Pacific Northwest. Maximum likelihood methods are used to constrain a
prior model for transient displacements in this region. The temporal
covariance of our prior model is described by a compact Wendland
covariance function, which significantly reduces the computational
burden that can be associated with GPR. Our results reveal the spatial
and temporal evolution of strain from slow slip events. We verify that
the transient strain estimated with GPR is in fact geophysical signal
by comparing it to the seismic tremor that is associated with Pacific
Northwest slow slip events.

@@

In addition to detecting geophysical processes, the GNSS derived
transient strain rates can be used to better understand the data from
borehole strain meters (BSMs). The Plate Boundary Observatory contains
about forty BSMs in the Pacific Northwest, and it has been
demonstrated that BSMs are able to record transient geophysical events
such as SSEs \citep[e.g.,][]{Dragert2011}. However, there are
complications that prevent BSM data from being used quantitatively in
geophysical studies. One difficulty is that BSM data should be
calibrated with a well known strain source, such as diurnal and
semidiurnal tides \citep{Hart1996,Roeloffs2010,Hodgkinson2013}.
Unfortunately, the tidal forces at BSMs which record SSEs are strongly
influenced by local bodies of water such as the Straight of Juan de
Fuca, making it difficult to form a theoretical prediction of tidal
strains \citep{Roeloffs2010}. Another complication is that noise in
BSM data is not well understood. The noise consists, in part, of a
long-term decay resulting from the instrument equilibrating with the
surrounding rock \citep{Gladwin1987}. Typically, this noise is dealt
with in an ad-hoc manner by fitting and removing exponentials and
low-order polynomials. We envision that the GNSS derived strain rates
from this paper can be used as a reference strain for calibrating BSM
data and quantify its noise.

%% CONCLUSION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}\label{sec:Conclusion}

We propose using Gaussian process regression (GPR) to estimate
transient strain rates from GNSS data. Most other methods for
estimating strain rates assume a parametric representation of
deformation, which can bias the results if the parameterization is not
chosen carefully. Here we assume a stochastic, rather than parametric,
prior model for displacements. Our prior model describes how much we
expect transient displacements to covary spatially and temporally. If
we know nothing about the underlying signal that we are trying to
recover, then the prior model can be chosen objectively with maximum
likelihood methods. We demonstrate that GPR is an effective tool for
detecting geophysical processes, such as slow slip events, in our
application to the Pacific Northwest. While this paper just focuses on
using GPR to estimate transient strain rates and detect geophysical
processes, we believe that GPR is a powerful tool that can be applied
to a wide range of geophysical problems.
